Perturbing SU($n$)-symmetric interactions

### 2020-01-23 collective_sun/sun_perturbation.tex
consider the case of an inhomogeneous two-body perturbation in the TI analytics

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-body
operator $O_{pq}$.  If the two-body operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ indexed by a scalar $\Delta$ is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
Solving the general eigenvalue problem in \eqref{eq:cond_2} requires
diagonalizing the ${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$
matrix $\check{\m h}$.  When the interaction Hamiltonian $H_0$ is
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

@@

Here we simplify the eigenvalue problems that appear at second order
in the single- and two-body theory in Sections
\ref{sec:single_body_pert} and \ref{sec:two_body_pert} in the case of
a translationally invariant system with periodic boundary conditions.
We solve the single-body eigenvalue problem analytically, and reduce
the two-body eigenvalue problem to that of diagonalizing an
$\p{N-1}\times\p{N-1}$ matrix.  Finally, for we discuss the further
reduction of the two-body problem to that of diagonalizing a matrix
with dimensions $\sim\p{N/2^D}\times\p{N/2^D}$ in the case of an
isotropic system on a $D$-dimensional periodic lattice, and we work
out the case of $D=1$ explicitly as an example.

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \p{2h_{pq}-h_p-h_q} w_{pq}
  + \sum_{k\notin\set{p,q}} \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
where $h_p\equiv\sum_q h_{pq}$.  If the interaction Hamiltonian $H_0$
and perturbation $\V_2\p{\m w}$ are translationally invariant, then
the couplings $h_{pq}$ and coefficients $w_{pq}$ depend on $p$ and $q$
only through the separation $\pm\p{p-q}$.  In this case, all $h_p=h_0$
are equal, and \eqref{eq:cond_2_ref_start} can be written in the form
\begin{align}
  2 \m h\circ \m w + \m h\c\m w + \m w\c\m h
  = \sp{\Delta + 2h_0} \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$ and $\m w$ are matrices of all $h_{pq}$ and $w_{pq}$, and
$\m h\circ\m w$ denotes an element-wise (Kronecker) product of $\m h$
and $\m w$, i.e.~with matrix elements $h_{pq}w_{pq}$.  In order to
find solutions to \eqref{eq:cond_2_ref}, we we expand the matrices
$\m h,\m w$ in a basis that respects translational symmetry.  To this
end, for all lattice site displacements $d\in\tilde\ZZ_L^D$ we define
the {\it shift matrices}
\begin{align}
  \m s_d \equiv \sum_p \op{p+d}{p},
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  The shift
matrices allow us to expand
\begin{align}
  \m h = \sum_{p,q} h_{pq} \op{p}{q}
  = \sum_{d,q} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m s_d,
  &&
  \m w = \sum_d w_{d,0} \m s_d,
\end{align}
and write the condition in \eqref{eq:cond_2_ref} as
\begin{align}
  \sum_{d\ne 0} h_{d,0} w_{d,0} \m s_d
  + \sum_{\substack{c,d\ne0\\c+d\ne0}} h_{c,0} w_{d,0} \m s_{c+d}
  = \p{\f12\Delta + h_0} \sum_{d\ne 0} w_{d,0} \m s_d,
\end{align}
where we explicitly exclude terms that always vanish due to the fact
that $\m h,\m w$ are zero on the diagonal.  With some re-indexing, we
can write
\begin{align}
  \sum_{d\ne 0} h_{d,0} w_{d,0} \m s_d
  + \sum_{\substack{c,d\ne0}} h_{d-c,0} w_{c,0} \m s_d
  = \p{\f12\Delta + h_0} \sum_{d\ne 0} w_{d,0} \m s_d,
  \label{eq:cond_2_shift}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}
  \equiv \sum_{d\ne0} h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \op{d}{c},
  &&
  \ut{\v{\m w}} \equiv \sum_{d\ne0} w_{d,0} \ket{d},
\end{align}
in order to express \eqref{eq:cond_2_shift} as the eigenvalue
equation
\begin{align}
  \ut{\check{\m h}} \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.
  \label{eq:cond_2_square}
\end{align}
We can reduce this eigenvalue problem even further in the case of an
isotropic system, in which $h_{pq}$ and $w_{pq}$ depend only on the
distance $\abs{p-q}$.  On a $D$ dimensional lattice, isotropy implies
one reflection symmetry for each lattice axis, which implies that the
number of degrees of freedom in $\ut{\v{\m w}}$ gets reduced by a
factor of $\sim 2^D$.  Isotropy similarly reduces the degrees of
freedom in $\ut{\check{\m h}}$ from $\sim N\times N$ to
$\sim\p{N/2^D}\times\p{N/2^D}$.  We work out the case of $D=1$ below
as an example, and leave the general case of arbitrary dimension
$D\ge1$ to future work.

In order to exploit reflection symmetry in one dimension ($D=1$), we
define a complete basis of orthogonal vectors that are symmetric under
reflection: for all $d\in\VV_N\equiv\set{1,2,\cdots,\floor{N/2}}$,
\begin{align}
  \ket{d_+} \equiv \f{\eta_d}{\sqrt{2}}\p{\ket{d} + \ket{-d}},
  &&
  \eta_d \equiv
  \begin{cases}
    1 & d \ne N/2, \\
    1/\sqrt{2} & d = N/2
  \end{cases},
\end{align}
where the special case of $d=N/2$ affects only even $N$, for which
$\ket{N/2}=\ket{-N/2}$.  We can use these vectors to define a
projector onto the reflection-symmetric subspace and expand
$\ut{\v{\m w}}$ as
\begin{align}
  \P_+ \equiv \sum_{d\in\VV_N} \op{d_+},
  &&
  \ut{\v{\m w}} = \sum_{d\in\VV_N} w_{d,0} \,
  \eta_d \sqrt{2} \ket{d_+}.
\end{align}
Finally, we project $\ut{\check{\m h}}$ onto the reflection-symmetric
subspace,
\begin{align}
  \P_+ \ut{\check{\m h}} \P_+
  &= \sum_{a,b\in\VV_N} \sum_{c,d\ne0}
  \p{h_{d,0} \delta_{cd} + h_{c-d,0}} \op{a_+} \op{d} \op{b_+} \\
  &= \sum_{a,b\in\VV_N} \op{a_+}{b_+} \f12 \eta_a\eta_b
  \sum_{c,d\ne0} \p{h_{d,0} \delta_{cd} + h_{c-d,0}}
  \p{\delta_{ac} + \delta_{-a,c}} \p{\delta_{bd} + \delta_{-b,d}} \\
  &= \sum_{a\in\VV_N} h_{a,0} \op{a_+}
  + \sum_{a,b\in\VV_N} \eta_a\eta_b
  \p{h_{a-b,0} + h_{a+b,0}} \op{a_+}{b_+},
\end{align}
which allows us to write down the reduced eigenvalue problem
\begin{align}
  \P_+ \ut{\check{\m h}} \P_+ \c \ut{\v{\m w}}
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}.
\end{align}
Note that the factors of $\eta_a,\eta_b$ in
$\P_+ \ut{\check{\m h}} \P_+$ only matter when $N$ is even, in which
case these factors merely multiply the last row and last column of
$\P_+ \ut{\check{\m h}} \P_+$ by a factor of $1/\sqrt{2}$.

>>>>

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the fully
symmetric manifold $\M_0$ by the two-body perturbation $\V_2$ in
\eqref{eq:perturbations} is simply the restriction of $\V_2$ onto
$\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQFS \sum_O \sum_{p<q} w_{Opq} O_{0,1}
  \EQFS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-body
operator $O_{pq}$.  If the two-body operator $O=X\otimes Y$, then
\begin{align}
  H_2^{(1)} \EQFS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} - \col{XY}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary fully symmetric state
$\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V2_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \f12 \sum_{p\ne q} \sp{2\bar h_{pq} w_{Opq}
    + \sum_k \p{h_{pk} w_{Okq} + w_{Opk} h_{kq}}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_{pq} - \f12\p{h_p + h_q},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ indexed by a scalar $\Delta$ is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h} \equiv 2\sum_{p<q} \bar h_{pq} \op{\set{p,q}}
  + \sum_{p<q} \ket{\set{p,q}} \sum_{k\notin\set{p,q}}
  \p{h_{pk} \bra{\set{k,q}} + h_{kq} \bra{\set{p,k}}}.
  \label{eq:h_super_mat}
\end{align}
Solving the general eigenvalue problem in \eqref{eq:cond_2} requires
diagonalizing the ${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$
matrix $\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

@@

Here we simplify the eigenvalue problems that appear at second order
in the single- and two-body theory in Sections
\ref{sec:single_body_pert} and \ref{sec:two_body_pert} in the case of
a translationally invariant system with periodic boundary conditions.
We solve the single-body eigenvalue problem analytically, and reduce
the two-body eigenvalue problem to that of diagonalizing
$\p{N-1}\times\p{N-1}$ matrices.  Finally, for we discuss the further
reduction of the two-body problem to that of diagonalizing matrices
with dimensions $\sim\p{N/2^D}\times\p{N/2^D}$ in the case of an
isotropic system on a $D$-dimensional periodic lattice, but leave this
reduction to future work.

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \p{2h_{pq}-h_p-h_q} w_{pq}
  + \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \Delta w_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
where $h_p\equiv\sum_q h_{pq}$.  If the interaction Hamiltonian $H_0$
is translationally invariant, then the couplings $h_{pq}$ depend on
$p$ and $q$ only through the separation $\pm\p{p-q}$.  In this case,
all $h_p=h_0$ are equal, and \eqref{eq:cond_2_ref_start} can be
written in the form
\begin{align}
  2 \m h\circ \m w + \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h}
  = \p{\Delta + 2h_0} \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$ and $\m w$ are matrices of all $h_{pq}$ and $w_{pq}$, and
$\m h\circ\m w$ denotes an element-wise (Kronecker) product of $\m h$
and $\m w$, and
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices.  In order to find solutions to \eqref{eq:cond_2_ref}, we we
expand the matrices $\m h,\m w$ in a basis that the respects
translational symmetry of $\m h$.  To this end, for all lattice
displacements $d\in\tilde\ZZ_L^D$ and wavenumbers
$k\in\tilde\ZZ_L^D\times2\pi/L$ we define the {\it cycle matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms of the cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
expand
\begin{align}
  \m h = \sum_{p,q} h_{pq} \op{p}{q}
  = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0},
  &&
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
\end{align}
and write the condition in \eqref{eq:cond_2_ref} as a new set of
conditions
\begin{align}
  \sum_{d\ne0} 2 h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  = \p{\Delta + 2 h_0} \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{d\ne0} 2 h_{d,0} \tilde w_{dk} \m m_{dk}
  + \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  = \p{\Delta + 2 h_0} \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{d\ne0} 2 h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k
  = \p{\Delta + 2 h_0} \ut{\v{\m w}}_k.
  \label{eq:cond_2_square}
\end{align}
The appearance of a wavenumber index $k$ in \eqref{eq:cond_2_square}
implies that all fixed-$k$ components of a general coefficient matrix
$\m w$ must simultaneously satisfy \eqref{eq:cond_2_square} in order
to generate states of fixed excitation energy $\Delta$ when the
corresponding perturbation $\V_2^\Delta$ acts on a fully symmetric
state $\ket\psi\in\M_0$.  The decomposition of an arbitrary
coefficient vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into
coefficient vectors $\v{\m w}_\Delta$ that generate a fixed-energy
perturbation $\V_2^\Delta$ is then performed by projecting each
$k$-component of $\v{\m w}$ onto the corresponding $\Delta$-eigenspace
of $\ut{\check{\m h}}_k$.  If the perturbation $\V_2$ is
translationally invariant, then the $k$-components of $\v{\m w}$ for
all $k\ne 0$ all vanish, so we only need to consider the case of
$k=0$.

We can reduce the eigenvalue problem in \eqref{eq:cond_2_square} even
further in the case of an isotropic system, in which the couplings
$h_{pq}$ depend only on the distance $\abs{p-q}$.  On a $D$
dimensional lattice, isotropy implies one reflection symmetry for each
lattice axis, which reduces the degrees of freedom in of
$\ut{\check{\m h}}_k$ from $\sim N\times N$ to
$\sim\p{N/2^D}\times\p{N/2^D}$.  We leave this reduction to future
work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The fully collective model}

[TODO: explicitly work out the special case of a fully collective
model]
